---
title: "The Grounded AI Advisor: A Dieselgate Walkthrough"
author: "Simon Mueller"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Grounded AI Advisor: A Dieselgate Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
# Keep this vignette offline and API-key-free at build time. The deterministic
# advisor chunks below evaluate live against the bundled `dieselgate` data; the
# LLM chunk is eval = FALSE and its output is shown as a static captured block.
Sys.setenv(EVENTSTUDY_NO_NETWORK = "1")
```

## Why the advisor exists

Event study analysis produces a lot of numbers — abnormal returns, cumulative
abnormal returns, t-statistics, normality p-values, autocorrelation
diagnostics — and interpreting them correctly requires methodological judgment
that is easy to get wrong. A large language model (LLM) is a tempting tool for
that interpretation, but a naive LLM will happily *fabricate* a plausible
number, cite a t-statistic the analysis never computed, or recommend a test on
the basis of a diagnostic that does not exist.

EventStudy's advisor is built so that this cannot happen. Its governing rule is
a single **grounding invariant**:

> The advisor never fabricates a number. Every claim it returns is provably tied
> to a diagnostic the package actually computed.

## How it works: two layers

The advisor is deliberately split into two layers.

1. **The deterministic offline grounding layer.** This is pure R with no network
   access, no API key, and no optional dependencies required. It harvests the
   already-computed statistical signals from a fitted task
   (`es_diagnostics()`), matches them against a curated, peer-reviewed knowledge
   base of methodology rules, and returns rule-based advice
   (`recommend_stat()`, `flag_robustness()`). This layer is always available and
   always reproducible.

2. **The LLM interpretation layer.** This is optional. When you supply a provider
   (`es_advise(..., provider = provider("anthropic"))`), the LLM is given *only*
   the package-computed diagnostics and the pre-grounded knowledge-base
   recommendations, and it is asked to write the prose interpretation around
   them — not to invent new numbers.

The grounding invariant is enforced by a **runtime guard** in R, independent of
the prompt. After the LLM responds, every recommendation's cited evidence is
re-checked against the actual diagnostics: any recommendation whose evidence
cites a diagnostic key that is absent, or a value that mismatches beyond a
numerical tolerance, is dropped, one caveat is recorded, and exactly one warning
is emitted. The LLM cannot talk its way past the guard because the guard does
not read the prompt — it reads the numbers.

This vignette walks through both layers on a real dataset. **The deterministic
layer runs live**; the **LLM layer is shown as a captured static example**,
because CRAN builds run offline with no API key (see the note at the end).

## The dieselgate example

The package bundles a small, frozen dataset, `dieselgate`, containing daily
adjusted prices for Volkswagen AG (`VOW.DE`, Xetra) and the DAX benchmark index
(`^GDAXI`) around the 2015 emissions scandal. On 2015-09-18 the U.S.
Environmental Protection Agency issued its Notice of Violation to Volkswagen;
the share-price crash lands on the following trading days.

`dieselgate` is a named list carrying three tibbles that feed
`EventStudyTask$new()` — `firm`, `index`, and a one-row `request` — plus
provenance `meta`. See `?dieselgate` for the full schema and provenance.

```{r load-data}
library(EventStudy)
data(dieselgate)

names(dieselgate)
dieselgate$request
```

We build the task from the three tibbles and run the standard pipeline
(`prepare_event_study()` -> `fit_model()` -> `calculate_statistics()`, wrapped by
`run_event_study()`) with a plain market model.

```{r run-pipeline}
task <- EventStudyTask$new(
  dieselgate$firm,
  dieselgate$index,
  dieselgate$request
)
task <- run_event_study(task, ParameterSet$new())
```

The market model estimated on the pre-event window has a beta near 1.1 and an
R-squared around 0.70 — a well-behaved fit. The event window tells the story of
the crash:

```{r show-ar-car}
ar  <- task$get_ar(1L)
car <- task$get_car(1L)

# Abnormal returns on the event day (0) and the two days after
ar[ar$relative_index %in% c(0, 1, 2, 3), ]

# Cumulative abnormal return over the full [-10, +10] window
tail(car, 1L)
```

The first trading day after the disclosure shows an abnormal return of roughly
**-17%**, followed by roughly **-12%** the next day, and the cumulative abnormal
return over the `[-10, +10]` window reaches about **-35%**. This is a sharp,
economically enormous, and highly statistically significant crash — exactly the
kind of result the advisor should help a researcher interpret and stress-test.

### Seeing the crash

The numbers above are easier to read as pictures. The two plots below render
**offline** from the AR and CAR tibbles already computed above — no network, no
API key. The abnormal-return plot makes the crash on the days *after* the event
day (relative index 0, drawn in a contrasting colour) visually obvious; the
cumulative-abnormal-return path shows the steady slide to about -35%.

```{r plot-ar, fig.alt = "Abnormal returns across the [-10, +10] event window, with the largest negative bars on the first two days after the event.", fig.width = 7, fig.height = 4}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ar$is_event_day <- ar$relative_index == 0
  ggplot2::ggplot(
    ar,
    ggplot2::aes(x = relative_index, y = abnormal_returns, fill = is_event_day)
  ) +
    ggplot2::geom_col() +
    ggplot2::scale_fill_manual(
      values = c(`FALSE` = "grey40", `TRUE` = "firebrick"),
      guide  = "none"
    ) +
    ggplot2::geom_hline(yintercept = 0, colour = "black", linewidth = 0.3) +
    ggplot2::labs(
      title = "Abnormal returns around the Dieselgate disclosure",
      subtitle = "Event day (relative index 0) highlighted",
      x = "Trading days relative to event",
      y = "Abnormal return"
    ) +
    ggplot2::theme_minimal()
} else {
  # ggplot2 is a hard dependency, but stay resilient in stripped-down builds.
  barplot(
    ar$abnormal_returns,
    names.arg = ar$relative_index,
    col = ifelse(ar$relative_index == 0, "firebrick", "grey40"),
    main = "Abnormal returns around the Dieselgate disclosure",
    xlab = "Trading days relative to event",
    ylab = "Abnormal return"
  )
  abline(h = 0)
}
```

The event-day abnormal return itself is small; the damage lands on the **next
two trading days**, which is why event windows are drawn wide enough to capture
the reaction rather than just the announcement instant.

```{r plot-car, fig.alt = "Cumulative abnormal return path across the [-10, +10] window, sliding to roughly -35 percent after the event.", fig.width = 7, fig.height = 4}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(car, ggplot2::aes(x = relative_index, y = car)) +
    ggplot2::geom_line(colour = "firebrick", linewidth = 0.8) +
    ggplot2::geom_point(colour = "firebrick", size = 1.2) +
    ggplot2::geom_hline(yintercept = 0, colour = "black", linewidth = 0.3) +
    ggplot2::geom_vline(
      xintercept = 0, colour = "grey50", linetype = "dashed", linewidth = 0.3
    ) +
    ggplot2::labs(
      title = "Cumulative abnormal return (CAR) path",
      subtitle = "Dashed line marks the event day; CAR reaches about -35%",
      x = "Trading days relative to event",
      y = "Cumulative abnormal return"
    ) +
    ggplot2::theme_minimal()
} else {
  plot(
    car$relative_index, car$car,
    type = "b", col = "firebrick",
    main = "Cumulative abnormal return (CAR) path",
    xlab = "Trading days relative to event",
    ylab = "Cumulative abnormal return"
  )
  abline(h = 0)
  abline(v = 0, lty = 2, col = "grey50")
}
```

By the end of the window the cumulative abnormal return has reached roughly
-35% — the headline crash the advisor will help interpret and stress-test below.

## What runs offline (deterministic)

Everything in this section evaluates **live**, with **no API key and no network
access**. This is the always-available grounding layer.

### Diagnostics

`es_diagnostics()` harvests the already-computed signals from the fitted task
into a structured, JSON-ready object. It recomputes nothing.

```{r diagnostics}
diag <- es_diagnostics(task)
diag
```

Note the diagnostics that matter here: the estimation-window R-squared is high
(~0.70), but the Shapiro-Wilk normality p-value is very small (well below 0.05),
so the residuals are **non-normal** — and there is only **one** event, a
single-firm study.

### Recommended statistics

`recommend_stat()` matches the diagnostics against the knowledge base's
statistical-choice rules and returns severity-ranked, citation-backed advice —
deterministically.

```{r recommend-stat}
recommend_stat(diag)
```

Because the estimation-window residuals fail the normality test, the KB fires a
rule recommending **non-parametric alternatives** (Sign Test, Rank Test) that
are robust to departures from normality, citing Brown & Warner (1985). This
recommendation is grounded in the actual Shapiro-Wilk p-value the package
computed.

### Robustness flags

`flag_robustness()` matches the diagnostics against the robustness rules.

```{r flag-robustness}
flag_robustness(diag)
```

Here the KB flags the **small-sample** problem: with a single event, the
cross-sectional t-test's large-sample normal approximation is unreliable, so
parametric p-values should be interpreted cautiously. Again, this flag is tied
directly to the computed event count.

Both of these calls are pure R. They return the same answer on every machine,
every time, with no external service involved.

## What the LLM adds (shown static)

The deterministic layer above tells you *which* tests to consider and *why*, but
it does so in terse, rule-based prose. The optional LLM layer takes the **same
grounded diagnostics and pre-grounded recommendations** and writes a fuller,
narrative interpretation around them — without inventing any numbers.

You would invoke it like this:

```{r llm-call, eval = FALSE}
# Requires an LLM provider and, for Anthropic/OpenAI-compatible providers,
# the optional `httr2` + `jsonlite` Suggests and an API key in the environment.
p      <- provider("anthropic")                 # or provider("openai_compatible", ...)
advice <- es_advise(diag, task_type = "recommend_stat", provider = p)
print(advice)
```

The block below is a **captured, illustrative example** of what
`print(advice)` returns for this dieselgate result. It is **not executed at
build time** (see the note that follows). Every number it cites — beta ~1.09,
R-squared ~0.70, the Shapiro-Wilk p-value, the event-window CAR and its
t-statistic — is one the live deterministic layer above already produced; the
LLM only supplies the surrounding prose, and the runtime guard would drop any
recommendation citing a diagnostic that did not match.

<!-- BEGIN static captured LLM example (NOT executed) -->
```
Event Study Advice
==================
Source:         anthropic
Task type:      recommend_stat
Deterministic:  FALSE
Recommendations: 1

Interpretation:
  The market model is well specified on the pre-event window (beta approximately
  1.09, R-squared approximately 0.70), so the estimated normal-return benchmark
  is credible. Against that benchmark the disclosure produced an abnormal return
  of roughly -17% on the first post-event trading day and roughly -12% on the
  second, cumulating to a cumulative abnormal return of about -35% over the
  [-10, +10] window (CAR t-statistic approximately -7.7). The effect is large and
  unambiguous. However, the estimation-window residuals fail the Shapiro-Wilk
  normality test (p approximately 0.004), which undermines the distributional
  assumptions of the standard parametric AR/CAR t-tests.

[1] Consider non-parametric test statistics (Sign Test, Rank Test)
    Kind:   stat_choice
    Effect: Preserves test size when estimation-window residuals are non-normal.
    Evidence:
      estimation_window.shapiro_p = 0.0042 (threshold 0.05, below)

Caveats:
 - This is a single-firm, single-event study; the cross-sectional t-test's
   large-sample approximation is unreliable with one event, so lean on the
   event's own time-series standard error and on non-parametric tests.
 - The crash coincides with heavy news flow and possible confounding events
   (regulatory statements, credit-rating actions); attribute the full CAR to the
   disclosure only with that caveat in mind.

Advisor Pro (waitlist): richer, corpus-grounded interpretation.
```
<!-- END static captured LLM example -->

### Why this block is not evaluated at build time

This vignette is built by CRAN and by `R CMD build` in an environment that has
**no network access and no API key**, and the optional AI Suggests
(`httr2`, `jsonlite`) may not be installed. A live `es_advise()` call with a
provider would need all three. So the LLM chunk is marked `eval = FALSE` and its
output is shown as the static block above. The deterministic chunks
(`es_diagnostics()`, `recommend_stat()`, `flag_robustness()`) *do* run live,
because they are pure R and need nothing external.

This is the whole point of the two-layer design: the grounded, reproducible
layer is always available and always runs; the LLM layer is an optional
interpretation on top, and it is held to the same grounding invariant by a
runtime guard whether or not you can see it execute here.

## Summary

* The advisor **never fabricates a number** — the grounding invariant.
* The **deterministic layer** (`es_diagnostics()`, `recommend_stat()`,
  `flag_robustness()`) is pure, offline R and runs live everywhere.
* The **LLM layer** (`es_advise()` with a provider) writes prose around the
  same grounded diagnostics and is policed by a runtime guard.
* On the bundled `dieselgate` example, the pipeline recovers the ~-35% CAR
  crash, the deterministic layer recommends non-parametric tests (non-normal
  residuals) and flags the single-event small-sample caveat, and the LLM layer
  turns those grounded facts into a readable interpretation.

See `?es_diagnostics`, `?recommend_stat`, `?flag_robustness`, `?es_advise`, and
`?provider` for the full API, and `?dieselgate` for the dataset's provenance.
